Comparison of MCMC algorithms in Stochastic Volatility Models

### 2023-09-23 manuscript/main.tex
remove dot
added resarch context and intro.

\begin{abstract}.  
    The objective of this research is to compare the computational methods used to estimate Bayesian stochastic volatility models. A simulation study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo and their ability to sample from the model's posterior distribution. Specifically, Simulation Based Calibration (SBC) is used to check whether these sampling strategies are returning efficient and well calibrated posterior estimates. Key metrics of interest are the effective sample size to check the efficiency of the algorithm and tests of uniformity to assess the calibration of the posteriors. This will determine which algorithm is better at estimating stochastic volatility models based on the efficiency and accuracy of the sampling strategy.
\end{abstract}

@@

- Should make reference to workflow, and reparameterisation of SV.

\section{Background and Literature Review}

    \subsection{Stochastic Volatility}
        Stochastic volatility models are used in financial econometrics to model the variance or volatility of returns in a financial instrument. The return of a financial asset often exhibits patterns in the variance which can be modelled to improve prediction and estimates of risk or tail events. Unlike other volatility models such as ARCH or GARCH, stochastic volatility models treat the variance as a random variable. Furthermore, these models have a state space representation since the variance is expressed as a latent stochastic process. State space models can be difficult to estimate since there is a latent state for every data observation in the time series, which means there are often more parameters than data points. These models can be estimated using frequent frameworks but this research will focus on Bayesian estimation strategies.

        \citet{kim1998stochastic} estimate a discrete time, univariate stochastic volatility model using Bayesian methods which is outlined below. $y_t$ is the mean corrected returns of some asset for equally spaced intervals t. $\beta$ is a constant scaling factor which is also defined as $exp(\mu / 2)$ representing instantaneous volatility. $h_t$ is log volatility, where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the autoregressive parameter $\phi$ such that $|\phi|<1$. This autoregressive parameter represents the persistence or "stickiness" of log volatility and the dispersion parameter $\sigma_{\eta}$ is the constant variance of the states. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks and are uncorrelated with each other.

@@

    \subsection{Gaussian Mixture Approximation}
        Kim, Shephard and Chib (KSC) sample the posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings and the Kalman Filter and smoother\footnote{In this research the exact software to apply the simulation smoother is unavailable, so a more recent simulation smoother is used which is based on the software written by the same author.} \citep{dejong1995}.

    \subsection{Hamiltonian Monte Carlo}
        Since the paper was written, new MCMC algorithms have been developed and enabled the estimation of richer and more complicated models. Specifically, Hamiltonian Monte Carlo is a MCMC algorithm which has become widely available for efficiently sampling from sophisticated models. Hamiltonian Monte Carlo, originally called Hybrid Monte Carlo, was developed in the physics literature \citep{duane1987hybrid} before being applied in the statistics literature by Radford Neal through his works in Bayesian Neural Networks \citep{neal1995bayesian} and statistical computing \citep{neal2011mcmc}. The algorithm has since become widely available through open source development projects such as Stan \citep{stan} and PyMC \citep{pymc2023}.

@@

>>>>

\begin{abstract}
    The objective of this research is to compare the computational methods used to estimate Bayesian stochastic volatility models. A simulation study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo and their ability to sample from the model's posterior distribution. Specifically, Simulation Based Calibration (SBC) is used to check whether these sampling strategies are returning efficient and well calibrated posterior estimates. Key metrics of interest are the effective sample size to check the efficiency of the algorithm and tests of uniformity to assess the calibration of the posteriors. This will determine which algorithm is better at estimating stochastic volatility models based on the efficiency and accuracy of the sampling strategy.
\end{abstract}

@@

Stochastic volatility is a type of volatility model which explicitly treats the variance as a latent random variable. These are typically expressed as non linear Gaussian state space models and are difficult to estimate using classical methods. The likelihood is unavailable in closed form and there are at least as many variables as data points. Bayesian approaches to estimating these models rely on computational Markov Chain Monte Carlo (MCMC) techniques to sample from the target joint posterior of these high dimensional latent parameter spaces.

Kim, Shephard and Chib (1998) published a method for sampling this model using a combination of Kalman Filters, Gibbs, and Metropolis Hastings over 20 years ago.  Since then, there have been many advances in Bayesian, and more broadly, statistical computation to estimate complex, high dimensional state space models. In particular, the recent availability of Hamiltonian Monte Carlo, a variant of the Metropolis Hastings algorithm, is a advancement of MCMC for efficient sampling of complex probabilistic models. The adoption of such new techniques have been made widely available through the delveopment of various open source libraries such as the Stan programming language and the PyMC library. Other algorithms have also improved the speed of estimating high dimensional models using approximations of the posterior such as Integrated nested Laplace approximation and Approximate Bayesian Computation.

These conceptually and practically different techniques to estimate complex models did not exist 20 years ago (or perhaps more accurately, were not easily accessible or implemented 20 years ago). As the development of new algorithms rapidly increase, so does the need to develop new methods to assess their output. How do we develop our workflow to assesss new sampling algorithms? And how can we compare the comptuational strategies used to estimate increasingly complex models?

% - SV models and more broadly state space models are complex, high dimensional models which require computational tools to estimate them.
% - Modelling variance useful and important?

% - Advanced in Bayesian computation
%     + Conceptually and practically diff techniques to estimate complex models didn't exist not that long ago
%     + Performance metrics useful for checking fit on single dataset. However not much emphasis on checking performance between MCMC

% - A principled Bayesian workflow which usually requires checking of multiple componenets: prior, posterior, model choice. 
%     + Lots of different components which could contribute to an incorrect fit or error.

% - This research focuses on evaluating the computational tools used to esitmate a specific type of model.

% % One para
% - This research </contributes> examine the algos used to estimate SV model 
%     + Particular focus is on the computational tools used to estimate these models and whether they are returning what we expect.
%     + conditional on the same prior

%     - Additionally, check the sensitivity of the simulation and algortihms to the parameterisation of the model. Taking idea and advice from Catherine's paper about improvements in sampling performance.

% %  2nd 
% - Investigating two algorithms, HMC and GMA from KSC.

% % 3
% - Two samplings methods are used and are compared through simulation
%     + Specifically, the simulation designed which is taken from talts et al.

% % 4
% - Key metrics of interests are ESS for efficiency and tests of uniformity for calibration.

% - Goal is to show which algo and parameterisation is "better". Moreover, want to demonstarte that this part of a Bayesian workflow which can be applied more generally is important for checking that the model is returning the correct posteriros - and if not, where the problem may be occurring and why. Give a set of tools to evaluate algos and parametiersations (before real data is even considered). **Why is this important.**

% - Sections outlined below:

\section{Resarch Context}

\subsection{Stochastic Volatility}
    \citet{kim1998stochastic} estimate a discrete time, univariate stochastic volatility model using Bayesian methods which is outlined below. $y_t$ is the mean corrected returns of some asset for equally spaced intervals t. $\beta$ is a constant scaling factor which is also defined as $exp(\mu / 2)$ representing instantaneous volatility. $h_t$ is log volatility, where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the autoregressive parameter $\phi$ such that $|\phi|<1$. This autoregressive parameter represents the persistence or "stickiness" of log volatility and the dispersion parameter $\sigma_{\eta}$ is the constant variance of the states. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks and are uncorrelated with each other.

@@

\subsection{Diagnostic limitations on real data}
The stochastic volatility model can be estimated using a variety of sampling strategies and MCMC algorithms. A common set of metrics are used to check the convergence of the MCMC. This includes checking the effective sample size and the $\hat{R}$ calculations for estimated chains and parameters. These diagnostics check whether an MCMC has failed to converge onto the target posterior distribution (since we can never know with certainty that a sampler has converged).

These diagnostics are useful for identifying when Bayesian computation fail on real data. However, confounding issues may arrise when trying to diagnose the cause of computational problems. Real data is generated from an unknown data generating process. That is, the true parameter or model is unobservable. Therefore, failed diagnostic checks could arise from either incorrect model specification or issues in the sampling strategy (algorithm) or both.

Furthermore, different sampling strategies and MCMC tools could provide different estimates for the same model. Assuming no issues in computation, these diagnostic measures do not provide any evidence to which estimate is closer to the truth.

To illustrate this problem, suppose we estimate this model with two different sampling strategies on real data. The data are the demeaned log returns of the 2019-2023 S\&P 500 Index. Below are the marginal posterior estimates for the static parameters $\mu$, $\phi$ and $\sigma^2$ and corresponding summary statistics.

**Insert results here**

These particular estimates show the marginal posteriors of each parameter sharing similar shapes. $\phi$ and $\sigma^2$ possess slightly heavier right tails for SM1 compared to SM2. While in this particular example there is no major difference in the estimates, there is no certainty in whether SM1 or SM2 is consistently closer to the truth. There are other ways of model or algorithmic selection in this context (for example, out of sample predictive performance), but for parameter estimation it is unclear which one should be selected.

\subsection{Diagnostic limitations on a single simulation}

Diagnosing computational problems on real data is difficult due to confounding factors. A strategy around this is to evaluate a model and sampling strategy on simulated data. One approach is to simulate data from the proposed model using known parameters. Then, fit the model on the simulated data and see if the model can recover the true parameters. This gives us the benefit of knowing or defining the data generating process to be estimated. If the model and algorithm cannot adequately capture the true parameter, then we cannot be confident that it will provide reliable estimates on real data.

Furthermore, as discussed in cite"Bayesian workflow", fitting models on simulated data is the only way we can check inference on latent variables. This is critical in Stochastic Volatility since the underlying framework is a state space model with latent log volatility parameters. Latent variables are unobserved in real data and are only estimated in the context of the model. Simulation gives control over the data generation which governs the latent process and reveals what the model can infer about the latent variables.

While simulations enable us to check whether our models and algorithms can capture the true generating process, there are limitations to what can be learned from a single simulation. There is always a small probability that the true parameter is in the tails of any estimated posterior distribution. Or put differently, there is a 1\% chance the true parameter or any random draw exists outside a 99\% credible interval. cite"Talts et al" make the point that a single simulation does not provide sufficient information about the inference made by an algorithm. As discussed in their paper, a single simulation may conclude "that an incorrectly coded analysis worked as desired, while a correctly coded analysis failed".

The below results are from a single simulation where data is generated and sampled from the stochastic volatility model with known parameters.

**Insert results here**

The posteriors for x and y sufficiently cover the true parameter, however, the true parameter for z is deep in the tails. Such an analysis may incorrectly conclude that the model fails to adequately estimate the z parameter. However, it may be the case that the posterior distribution is correctly calculated using this algorithm and the results may be due to the nuances of this specific simulated dataset.

\subsection{Research Goal}
The objective of this research is to design a simulation study to evaluate the calibration of algorithms used to sample Stochastic Volatility models. Indeed, there are limitations to evaluating MCMC algorithms based on fits to real data and single simulations. To check the calibration of an algorithm, simulations over multiple "true parameters" are required. This methodology is discussed in the next session.

Additionally, two sampling strategies will be assessed using the proposed simulation strategy as well as the performance of the samplers under different parameterisations of the stochastic volatility model. Not only will the calibration of any one algorithm be assessed, but also compared across sampling strategies to determine which MCMC approach is most suitable for estimating this model.

@@

\section{MCMC Strategies}

    \subsection{Sampling method 1}
        Kim, Shephard and Chib (KSC) sample the posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings and the Kalman Filter and smoother\footnote{In this research the exact software to apply the simulation smoother is unavailable, so a more recent simulation smoother is used which is based on the software written by the same author.} \citep{dejong1995}.

    \subsection{Sampling method 2}
    % Need much more detail here about how it works
        Since the paper was written, new MCMC algorithms have been developed and enabled the estimation of richer and more complicated models. Specifically, Hamiltonian Monte Carlo is a MCMC algorithm which has become widely available for efficiently sampling from sophisticated models. Hamiltonian Monte Carlo, originally called Hybrid Monte Carlo, was developed in the physics literature \citep{duane1987hybrid} before being applied in the statistics literature by Radford Neal through his works in Bayesian Neural Networks \citep{neal1995bayesian} and statistical computing \citep{neal2011mcmc}. The algorithm has since become widely available through open source development projects such as Stan \citep{stan} and PyMC \citep{pymc2023}.

### 2023-09-23 manuscript/structure/draft.qmd
added resarch context and intro.

\documentclass[12pt, a4paper]{article}
\setcounter{tocdepth}{2}
\usepackage{graphicx} % Required for inserting images
\usepackage[parfill]{parskip} % Removes indent from new paragraphs
\usepackage{amsmath, amssymb, amsfonts}
\usepackage[colorlinks = true,
			citecolor  = blue]{hyperref}

\usepackage[longnamesfirst,authoryear, round]{natbib}
\bibliographystyle{unsrtnat}

\usepackage[
top    = 1.0in,
bottom = 1.2in,
left   = 1.0in,
right  = 1.0in]{geometry}
\linespread{1.5}

\title{Comparison of MCMC algorithms in Stochastic Volatility Models}
\author{Benjamin Wee}
\date{October 2023}

\begin{document}

\maketitle

\begin{abstract}
    The objective of this research is to compare the computational methods used to estimate Bayesian stochastic volatility models. A simulation study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo and their ability to sample from the model's posterior distribution. Specifically, Simulation Based Calibration (SBC) is used to check whether these sampling strategies are returning efficient and well calibrated posterior estimates. Key metrics of interest are the effective sample size to check the efficiency of the algorithm and tests of uniformity to assess the calibration of the posteriors. This will determine which algorithm is better at estimating stochastic volatility models based on the efficiency and accuracy of the sampling strategy.
\end{abstract}

\newpage

\tableofcontents{\protect\newpage}

\section{Introduction}

Stochastic volatility is a type of volatility model which explicitly treats the variance as a latent random variable. These are typically expressed as non linear Gaussian state space models and are difficult to estimate using classical methods. The likelihood is unavailable in closed form and there are at least as many variables as data points. Bayesian approaches to estimating these models rely on computational Markov Chain Monte Carlo (MCMC) techniques to sample from the target joint posterior of these high dimensional latent parameter spaces.

@@

These conceptually and practically different techniques to estimate complex models did not exist 20 years ago (or perhaps more accurately, were not easily accessible or implemented 20 years ago). As the development of new algorithms rapidly increase, so does the need to develop new methods to assess their output. How do we develop our workflow to assesss new sampling algorithms? And how can we compare the comptuational strategies used to estimate increasingly complex models?

% - SV models and more broadly state space models are complex, high dimensional models which require computational tools to estimate them.
% - Modelling variance useful and important?

% - Advanced in Bayesian computation
%     + Conceptually and practically diff techniques to estimate complex models didn't exist not that long ago
%     + Performance metrics useful for checking fit on single dataset. However not much emphasis on checking performance between MCMC

% - A principled Bayesian workflow which usually requires checking of multiple componenets: prior, posterior, model choice. 
%     + Lots of different components which could contribute to an incorrect fit or error.

% - This research focuses on evaluating the computational tools used to esitmate a specific type of model.

% % One para
% - This research </contributes> examine the algos used to estimate SV model 
%     + Particular focus is on the computational tools used to estimate these models and whether they are returning what we expect.
%     + conditional on the same prior

%     - Additionally, check the sensitivity of the simulation and algortihms to the parameterisation of the model. Taking idea and advice from Catherine's paper about improvements in sampling performance.

% %  2nd 
% - Investigating two algorithms, HMC and GMA from KSC.

% % 3
% - Two samplings methods are used and are compared through simulation
%     + Specifically, the simulation designed which is taken from talts et al.

% % 4
% - Key metrics of interests are ESS for efficiency and tests of uniformity for calibration.

% - Goal is to show which algo and parameterisation is "better". Moreover, want to demonstarte that this part of a Bayesian workflow which can be applied more generally is important for checking that the model is returning the correct posteriros - and if not, where the problem may be occurring and why. Give a set of tools to evaluate algos and parametiersations (before real data is even considered). **Why is this important.**

% - Sections outlined below:

\section{Resarch Context}

\subsection{Stochastic Volatility}
    \citet{kim1998stochastic} estimate a discrete time, univariate stochastic volatility model using Bayesian methods which is outlined below. $y_t$ is the mean corrected returns of some asset for equally spaced intervals t. $\beta$ is a constant scaling factor which is also defined as $exp(\mu / 2)$ representing instantaneous volatility. $h_t$ is log volatility, where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the autoregressive parameter $\phi$ such that $|\phi|<1$. This autoregressive parameter represents the persistence or "stickiness" of log volatility and the dispersion parameter $\sigma_{\eta}$ is the constant variance of the states. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks and are uncorrelated with each other.

@@

\subsection{Diagnostic limitations on real data}
The stochastic volatility model can be estimated using a variety of sampling strategies and MCMC algorithms. A common set of metrics are used to check the convergence of the MCMC. This includes checking the effective sample size and the $\hat{R}$ calculations for estimated chains and parameters. These diagnostics check whether an MCMC has failed to converge onto the target posterior distribution (since we can never know with certainty that a sampler has converged).

These diagnostics are useful for identifying when Bayesian computation fail on real data. However, confounding issues may arrise when trying to diagnose the cause of computational problems. Real data is generated from an unknown data generating process. That is, the true parameter or model is unobservable. Therefore, failed diagnostic checks could arise from either incorrect model specification or issues in the sampling strategy (algorithm) or both.

@@

To illustrate this problem, suppose we estimate this model with two different sampling strategies on real data. The data are the demeaned log returns of the 2019-2023 S\&P 500 Index. Below are the marginal posterior estimates for the static parameters $\mu$, $\phi$ and $\sigma^2$ and corresponding summary statistics.

@@

These particular estimates show the marginal posteriors of each parameter sharing similar shapes. $\phi$ and $\sigma^2$ possess slightly heavier right tails for SM1 compared to SM2. While in this particular example there is no major difference in the estimates, there is no certainty in whether SM1 or SM2 is consistently closer to the truth. There are other ways of model or algorithmic selection in this context (for example, out of sample predictive performance), but for parameter estimation it is unclear which one should be selected.

\subsection{Diagnostic limitations on a single simulation}

Diagnosing computational problems on real data is difficult due to confounding factors. A strategy around this is to evaluate a model and sampling strategy on simulated data. One approach is to simulate data from the proposed model using known parameters. Then, fit the model on the simulated data and see if the model can recover the true parameters. This gives us the benefit of knowing or defining the data generating process to be estimated. If the model and algorithm cannot adequately capture the true parameter, then we cannot be confident that it will provide reliable estimates on real data.

Furthermore, as discussed in cite"Bayesian workflow", fitting models on simulated data is the only way we can check inference on latent variables. This is critical in Stochastic Volatility since the underlying framework is a state space model with latent log volatility parameters. Latent variables are unobserved in real data and are only estimated in the context of the model. Simulation gives control over the data generation which governs the latent process and reveals what the model can infer about the latent variables.

While simulations enable us to check whether our models and algorithms can capture the true generating process, there are limitations to what can be learned from a single simulation. There is always a small probability that the true parameter is in the tails of any estimated posterior distribution. Or put differently, there is a 1\% chance the true parameter or any random draw exists outside a 99\% credible interval. cite"Talts et al" make the point that a single simulation does not provide sufficient information about the inference made by an algorithm. As discussed in their paper, a single simulation may conclude "that an incorrectly coded analysis worked as desired, while a correctly coded analysis failed".

@@

The posteriors for x and y sufficiently cover the true parameter, however, the true parameter for z is deep in the tails. Such an analysis may incorrectly conclude that the model fails to adequately estimate the z parameter. However, it may be the case that the posterior distribution is correctly calculated using this algorithm and the results may be due to the nuances of this specific simulated dataset.

\subsection{Research Goal}
The objective of this research is to design a simulation study to evaluate the calibration of algorithms used to sample Stochastic Volatility models. Indeed, there are limitations to evaluating MCMC algorithms based on fits to real data and single simulations. To check the calibration of an algorithm, simulations over multiple "true parameters" are required. This methodology is discussed in the next session.

@@

\section{Methodology}

    \subsection{Simulation Design}
        Simulation Based Calibration (SBC) is a technique that checks the calibration of posterior estimates generated by MCMC algorithms. SBC is conducted by comparing the distribution of rank statistics to the discrete uniform distribution which arises when an algorithm is correctly calibrated. The procedure starts by taking draws from the prior distribution and creating datasets implied by each draw. Rank statistics are then calculated on the posterior samples conditional on the simulated data.

@@

    \subsection{Metrics}
        The key metrics and diagnostics to compare the performance of these methods are the effective sample size (ESS), rank statistics, and chi-squared test statistics.

@@

\section{MCMC Strategies}

    \subsection{Sampling method 1}
        Kim, Shephard and Chib (KSC) sample the posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings and the Kalman Filter and smoother\footnote{In this research the exact software to apply the simulation smoother is unavailable, so a more recent simulation smoother is used which is based on the software written by the same author.} \citep{dejong1995}.

@@

    \subsection{Sampling method 2}
    % Need much more detail here about how it works
        Since the paper was written, new MCMC algorithms have been developed and enabled the estimation of richer and more complicated models. Specifically, Hamiltonian Monte Carlo is a MCMC algorithm which has become widely available for efficiently sampling from sophisticated models. Hamiltonian Monte Carlo, originally called Hybrid Monte Carlo, was developed in the physics literature \citep{duane1987hybrid} before being applied in the statistics literature by Radford Neal through his works in Bayesian Neural Networks \citep{neal1995bayesian} and statistical computing \citep{neal2011mcmc}. The algorithm has since become widely available through open source development projects such as Stan \citep{stan} and PyMC \citep{pymc2023}.

@@

    \subsection{Model Parameterisation}

        The parameterisation of a model can affect the performance of a MCMC algorithm when sampling from models with complex posterior geometries. An example of this is Neal's funnel \citep{neal2003slice} where the Hamiltonian Monte Carlo sampler encounters performance issues in hierarchical models and produces biased samples.

@@

        \subsubsection{Non centered Hamiltonian Monte Carlo}
        First sample from a standard normal distribuition and multiply by the variance of the log volatility.

@@

        \subsubsection{Non centered Gaussian Mixture}
        The non centered model for the Gaussian approximation is expressed slightly differently due to the use of the Kalman Filter. While HMC can sample from the joint posterior directly with state vectors centered at zero, applying the Kalman Filter requires the state equation to be a function of the state variable.

@@

        \subsubsection{Prior Predictive check}
        It is worth noting that the centered and non-centered parameteristions are the same stochastic volatility model. Different parameteristions express the same mathematical models in ways that may make it easier or harder for a given algorithm to sample.

@@

\section{Results}

    \subsection{Hamiltonian Monte Carlo}

    \subsection{Gaussian Mixture Approximation}

    \subsection{Reweigthing rank statistics of Gaussian Mixture}

\section{Discussion}

\subsection{Limitations}

\section{Conclusion}

\newpage

% \bibliographystyle{apa} % Econometrica referencing
\bibliography{references}

\end{document}

>>>>

# Abstract
The objective of this research is to compare the computational methods used to estimate Bayesian stochastic volatility models. A simulation study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo and their ability to sample from the model's posterior distribution. Specifically, Simulation Based Calibration (SBC) is used to check whether these sampling strategies are returning efficient and well calibrated posterior estimates. Key metrics of interest are the effective sample size to check the efficiency of the algorithm and tests of uniformity to assess the calibration of the posteriors. This will determine which algorithm is better at estimating stochastic volatility models based on the efficiency and accuracy of the sampling strategy.

# Introduction
Stochastic volatility is a type of volatility model which explicitly treats the variance as a latent random variable. These are typically expressed as non linear Gaussian state space models and are difficult to estimate using classical methods. The likelihood is unavailable in closed form and there are at least as many variables as data points. Bayesian approaches to estimating these models rely on computational Markov Chain Monte Carlo (MCMC) techniques to sample from the target joint posterior of these high dimensional latent parameter spaces.

@@

These conceptually and practically different techniques to estimate complex models did not exist 20 years ago (or perhaps more accurately, were not easily accessible or implemented 20 years ago). As the development of new algorithms rapidly increase, so does the need to develop new methods to assess their output. Developments in our statistical workfow are required to assess new sampling algorithms as well as compare computational strategies used to estimate increasingly complex models.

This research implements a simulation study to assess the algorithms used to estimate stochastic volatility models. Simulation Based Calibration outlined in Talts et al. is the primary study design to determine whether the MCMC strategy are returning efficient and well calibrated posterior estimates.

Two sampling approaches are considered in this study - Kalman Filter and Metropolis Hastings estimation by Kim Sherphard and Chib, and the Hamiltoninan Monte Carlo algorithim.

Key metrics used to evaluate the model are the effective sample size and visual tests of uniform rank statistics.

<!-- 
 How do we develop our workflow to assesss new sampling algorithms? And how can we compare the comptuational strategies used to estimate increasingly complex models? -->

<!-- How can we check whether these algorithms are returning the correct posterior estimates?  -->
are complex, high dimensional models that can be expressed in the state space framework.

A variety of MCMC strategies exist in the literature for estimating these models such as particle filters, Kalman Filters and Metropolis Hastings algorithms.

and more broadly state space models are complex, high dimensional models which require computational tools to estimate them.
- Modelling variance useful and important?

- Advanced in Bayesian computation
    + Conceptually and practically diff techniques to estimate complex models didn't exist not that long ago
    + Performance metrics useful for checking fit on single dataset. However not much emphasis on checking performance between MCMC

- A principled Bayesian workflow which usually requires checking of multiple componenets: prior, posterior, model choice. 
    + Lots of different components which could contribute to an incorrect fit or error.

- This research focuses on evaluating the computational tools used to esitmate a specific type of model.

% One para
- This research </contributes> examine the algos used to estimate SV model 
    + Particular focus is on the computational tools used to estimate these models and whether they are returning what we expect.
    + conditional on the same prior

    - Additionally, check the sensitivity of the simulation and algortihms to the parameterisation of the model. Taking idea and advice from Catherine's paper about improvements in sampling performance.

%  2nd 
- Investigating two algorithms, HMC and GMA from KSC.

% 3
- Two samplings methods are used and are compared through simulation
    + Specifically, the simulation designed which is taken from talts et al.

% 4
- Key metrics of interests are ESS for efficiency and tests of uniformity for calibration.

- Goal is to show which algo and parameterisation is "better". Moreover, want to demonstarte that this part of a Bayesian workflow which can be applied more generally is important for checking that the model is returning the correct posteriros - and if not, where the problem may be occurring and why. Give a set of tools to evaluate algos and parametiersations (before real data is even considered). **Why is this important.**

- Sections outlined below:

# Stochastic Volatility
The model of interest in this research is the discrete time, univariate stochastic volatility model outlined by \citet{kim1998stochastic} who use Bayesian estimation methods. $y_t$ is the mean corrected returns of some asset for equally spaced intervals t. $\beta$ is a constant scaling factor which is also defined as $exp(\mu / 2)$ representing instantaneous volatility. $h_t$ is log volatility, where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the autoregressive parameter $\phi$ such that $|\phi|<1$. This autoregressive parameter represents the persistence or "stickiness" of log volatility and the dispersion parameter $\sigma_{\eta}$ is the constant variance of the states. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks and are uncorrelated with each other.

@@

## Diagnostic limitations on real data
The stochastic volatility model can be estimated using a variety of sampling strategies and MCMC algorithms. A common set of metrics are used to check the convergence of the MCMC. This includes checking the effective sample size and the $\hat{R}$ calculations for estimated chains and parameters. These diagnostics check whether an MCMC has failed to converge onto the target posterior distribution (since we can never know with certainty that a sampler has converged).

Convergence diagnostics are useful for identifying when Bayesian computation fail on real data. However, confounding issues may arise when trying to diagnose the _cause_ of computational problems. Real data is generated from an unknown data generating process. That is, the true parameter or model is unobservable. Therefore, failed diagnostic checks could arise from either inocrrect model specification or issues in the sampling strategy (algorithm) or both.

@@

To illustrate this problem, suppose we estimate this model with two different sampling strategies on real data. The data are the demaned log returns of the 2019-2023 S&P 500 Index. Below are the marginal posterior estimates for the static parameters $\mu$, $\phi$ and $\sigma^2$ and corresponding summary statistics.

@@

These particular estimates show the marginal posteriors of each parameter sharing similar shapes. $\phi$ and $\sigma^2$ possess slightly heavier right tails for SM1 compared to SM2. While in this particular example there is no major difference in the estimates, there is no certainty in whether SM1 or SM2 is consistenty closer to the truth. There are other ways of model or algorithmic selection in this context (for example, out of sample predictive performance), but for parameter _estimation_ it is unclear which one should be selected.

<!-- Could simply just pick the one which has better out of sample performance. Although in the case when computation fails, we wouldn't -->
<!-- 
- Workflwo pg 16: "Modeling issues can collide with computational issues" -  can't tell if problem is computation or the model (and whether it's "correct" or capturing the correct DGP). -->

## Diagnostic limitations on a single simulation
Diagnosing computational problems on real data is difficult due to confounding factors. A strategy around this is to evaluate a model and sampling strategy on simulated data. One approach is to simulate data from the proposed model using known parameters. Then, fit the model on the simulated data and see if the model can recover the true parameters. This gives us the benefit of knowing or defining the data generating process to be estimated. If the model and algorithm cannot adequately capture the true parameter, then we cannot be confident that it will provide reliable estimates on real data.

Furthermore, as discussed in \cite{Bayesian workflow}, fitting models on simulated data is the only way we can check inference on latent variables. This is critical in Stochastic Volatility since the underlying framework is a state space model with latent log volatility parameters. Latent variables are unobserved in real data and are only estimated in the context of the model. Simulation gives control over the data generation which governs the latent process and reveals what the model can infer about the latent variables.

While simulations enable us to check whether our models and algorithms can capture the true generating process, there are limitations to what can be learned from a single simulation. There is always a small probability that the true parameter is in the tails of any estimated posterior distribution. Or put differently, there is a 1\% chance the true parameter or any random draw exists outside a 99\% credible interval. \cite{Talts et al} make the point that a single simulation does not provide sufficient information about the inference made by an algorithm. As discussed in their paper, a single simulation may conclude "that an incorrectly coded analysis worked as desired, while a correctly coded analysis failed".

@@

The posteriors for x and y sufficiently cover the true parameter, however, the true parameter for z is deep in the tails. Such an analysis may incorrectly conclude that the model fails to adequatly estimate the z parameter. However, it may be the case that the posterior distribution is correctly calculated using this algorithm and the results may be due to the nuances of this specific simulated dataset.

<!-- Workflow:
- page 17: simulate data from model with fixed known parameers and see whether method comes close to reproducing known truth. If the model cannot sufficiently capture the truth when the truth is known, then we cannot rely on estimates made on real data.
- pg 18 "Fake-data simulations provide an upper bound of what can be learned about a latent process" [since this is unobserved] 
Multiple ground truths needed.
-->
<!-- 
- pg 18: Bayes inferencconly calibrated when averaging over the prior - not for any one parameter alue (single simulation anything can happen, i.e a 5\% chance a random draw outside the 95\% interval)
SBC:
- pg 2:The behaviour of any individual siulation will not characterise the bility of the inference to fit the model in any meaningful way. May prematurely conclude that an incorrectly coded analysis works as desired or a correctly coded analysis failed. Need to consider multiple ground truths  -->

## Research Goal
The objective of this research is to design a simulation study to evaluate the calibration of algorithms used to sample Stochastic Volatility models. Indeed, there are limitations to evaluating MCMC algorithms based on fits to real data and single simulations. To check the calibration of an algorithm, simulations over multiple "true parameters" are required. This methodology is discussed in the next session.

@@

<!-- 
- Need to consider multiple ground truths
- Apply SBC to these 2 algorithms to assess their calibration. Tool allows us to not only assess calibration for a single algorithm but enables comparison between strategies (Not only get to validate any one algo but get to compare across algo.) -->

# Methodology

## Simulation Design
Simulation Based Calibration (SBC) checks the calibration of posterior estimates generated by MCMC algorithms. SBC is conducted by comparing the distribution of rank statistics to the discrete uniform distribution which arises when an algorithm is correctly calibrated. The procedure starts by taking draws from the prior distribution and creating datasets implied by each draw. Rank statistics are then calculated on the posterior samples conditional on the simulated data.

@@

## Metrics
The key metrics and diagnostics to compare the performance of these methods are the effective sample size (ESS), rank statistics, and chi-squared test statistics.

@@

# MCMC Strategies

## Sampling method 1
Kim, Shephard and Chib (KSC) sample the posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings and the Kalman Filter and smoother\footnote{In this research the exact software to apply the simulation smoother is unavailable, so a more recent simulation smoother is used which is based on the software written by the same author.} \citep{dejong1995}.

@@

## Sampling method 2
    % Need much more detail here about how it works
Since the paper was written, new MCMC algorithms have been developed and enabled the estimation of richer and more complicated models. Specifically, Hamiltonian Monte Carlo is a MCMC algorithm which has become widely available for efficiently sampling from sophisticated models. Hamiltonian Monte Carlo, originally called Hybrid Monte Carlo, was developed in the physics literature \citep{duane1987hybrid} before being applied in the statistics literature by Radford Neal through his works in Bayesian Neural Networks \citep{neal1995bayesian} and statistical computing \citep{neal2011mcmc}. The algorithm has since become widely available through open source development projects such as Stan \citep{stan} and PyMC \citep{pymc2023}.

@@

## Model Parameterisation
The parameterisation of a model can affect the performance of a MCMC algorithm when sampling from models with complex posterior geometries. An example of this is Neal's funnel \citep{neal2003slice} where the Hamiltonian Monte Carlo sampler encounters performance issues in hierarchical models and produces biased samples.

@@

### Non centered Hamiltonian Monte Carlo
First sample from a standard normal distribuition and multiply by the variance of the log volatility.

@@

### Non centered Gaussian Mixture
The non centered model for the Gaussian approximation is expressed slightly differently due to the use of the Kalman Filter. While HMC can sample from the joint posterior directly with state vectors centered at zero, applying the Kalman Filter requires the state equation to be a function of the state variable.

@@

### Prior Predictive check
It is worth noting that the centered and non-centered parameteristions are the same stochastic volatility model. Different parameteristions express the same mathematical models in ways that may make it easier or harder for a given algorithm to sample.

@@

# Results

## Hamiltonian Monte Carlo

## Gaussian Mixture Approximation

## Reweigthing rank statistics of Gaussian Mixture

# Discussion

## Limitations
- Weakness of SBC: 
    + Model uses very uninofmrative priors. Extremely unrealistic datasets which cause SBC to fail, which would otherwise work well on real data. 
    + Computationally expensive

# Conclusion
